\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
%\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{bbm}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{mathpartir}
\usepackage{mathabx}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{cprotect}
\usepackage{subcaption}

\usepackage{cleveref}
\AtBeginDocument{\renewcommand\ref\cref}
\crefname{figure}{Figure}{Figures}
\crefname{theorem}{Theorem}{Theorems}


\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

\input {macros}

\begin{document}

\lstset{language=ML,
        columns=fullflexible,
        basicstyle=\ttfamily,
        tabsize=4,
        escapeinside={"*}{*"},
        literate={|->}{{\ $\mapsto$\ }}1
                 {xhat}{{$\hat{\texttt{x}}$}}1
                 {yhat}{{$\hat{\texttt{y}}$}}1
                 {zhat}{{$\hat{\texttt{z}}$}}1
                 {murec}{{$\mu$}}1
                 {alpha}{{$\alpha$}}1
                 {beta}{{$\beta$}}1
                 {$}{{$\fut$}}1}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{20yy} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

% Uncomment one of the following two, if you are not going for the 
% traditional copyright transfer agreement.

%\exclusivelicense                % ACM gets exclusive license to publish, 
                                  % you retain copyright

%\permissiontopublish             % ACM gets nonexclusive license to publish
                                  % (paid open-access papers, 
                                  % short abstracts)

%\titlebanner{banner above paper title}        % These are ignored unless
%\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Stage-Splitting a Modal Language}

\authorinfo{Nicolas Feltman \and Carlo Angiuli \and Umut Acar \and Kayvon Fatahalian}
           {Carnegie Mellon University}
           {\{nfeltman,cangiuli,umut,kayvonf\}@cs.cmu.edu}

\maketitle

%\begin{abstract}
%\input{abstract}
%\end{abstract}

%\category{CR-number}{subcategory}{third-level}

% general terms are not compulsory anymore, 
% you may leave them out
%\terms
%term1, term2

%\keywords
%keyword1, keyword2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

It is a common observation that some computations can be performed efficiently by using {\em staging}, wherein the execution of a program is split over two or more discrete points in time ({\em i.e.} stages).  This most often arises when we want to evaluate a multivariate function several times, with some inputs dynamic (varying across invocations) and some static (constant across invocations).  Examples include,
\begin{itemize}
\item Computing $b^e$, where $b$ varies and $e$ is constant.
\item Finding the $i$th largest element in a list, where the list is constant and $i$ varies.
\item Computing the shading of a surface in a 3D rendering, where some contributors to the final pixel color can be approximated as constant over  pixel blocks.  
\end{itemize}

With these goals in mind, our initial concern is how to express multistage programs. To this end, we have created \lang, a typed multi-stage lambda calculus featuring products, sums, and isorecursive types. An important property of \lang\ is that the staging of a well-typed program is apparent without resorting to any variable analysis, contra standard binding time analysis.  

Once we have appropriately written {\em multi-stage} programs, we then must execute them in a way that respects the staging.  In this paper, we explore two techniques for this: {\em partial evaluation} and {\em stage-splitting}.  

The main idea of the former technique is to evaluate all of the constant portions of the program and embed the partial results into a {\em residual} program that depends only on the varying inputs.  This is an old and well-explored idea, though the typical setup of partial evaluation implies a more powerful binding time analysis than is necessary for our language.

The main idea of the latter technique is to statically compile the multi-stage program into several smaller programs, one per stage, and then to execute those in a standard way once inputs are available.  This has previously been explored as an optimization technique for languages with fewer features (notably, no functions) or a more constrained staging model than \lang. Prior examples include the ``data specialization" work of Knoblock and Ruff\,\cite{knoblock96} as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.  Unlike partial evaluation, stage splitting is purely a static program transformation and is not dependent on the input data.

\TODO motivate this example


\Cref{sec:semantics} describes the grammar and type system of \lang, as well as a reference semantics which is essentially a partial evaluator that respects staging. \Cref{sec:splitting} describes a stage-splitting transformation for \lang\ and proves its equivalence to the reference semantics.
Finally, \Cref{sec:examples} gives several examples of how stage-splitting terms in \lang\ can be used to derive classic data structures, and discusses the effectiveness and limitations of this technique.  Notably, these examples make use of the stage-splitting features not available in prior work.

\input{semantics}
\input{splitting}
\input{algoexamples}

%\appendix
%\section{Appendix Title}
%This is the text of the appendix, if you need one.

%\acks
%Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\bibliography{techreport}
\bibliographystyle{abbrvnat}

\end{document}
